Supporting Non-Functional Requirements in Services Software Development Process: an MDD Approach

### 2013-10-22 SOFSEM2014/FinalVersion/conclusion.tex
Final version to SOFSEM!

This paper presented the $\pi$SOD-M software method for specifying and designing service-based applications in the presence of  non-functional constraints. 
Our proposal enhances the SOD-M method with constraints, policies and contracts  to consider non-functional constraints of applications.
We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developping a simple application.

$\pi$SOD-M is being used in an academic environment.
So far, the preliminary results indicate that $\pi$SOD-M approach is useful for the development of complex web service applications. 
We are now working on the definition of a PSM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS).

>>>>

This paper presented the $\pi$SOD-M software method for specifying and designing service-based applications in the presence of  non-functional constraints. 
Our proposal enhances the SOD-M method with constraints, policies and contracts  to consider non-functional constraints of applications.
We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developing a simple application.

$\pi$SOD-M is being used in an academic environment.
So far, the preliminary results indicate that $\pi$SOD-M approach is useful for the development of complex web service applications. 
We are now working on the definition of a PSM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS) in order to be able to generate code for the de-facto service coordination standard.

### 2013-10-22 SOFSEM2014/FinalVersion/introduction.tex
Final version to SOFSEM!

Model Driven Development (MDD)~\cite{Favre06arigorous} is a top-down approach proposed by the Object Management Group
(OMG)\footnote{\texttt{http://www.omg.org/mda}.} for designing and developing software systems. 
MDD provides a set of guidelines for  structuring  specifications by using \textit{models} to specify software systems at different levels of abstraction or \textit{viewpoints}:\\[1mm]
\noindent 
-\textit{Computation Independent Models (CIM)}: this viewpoint represents the software system at its highest level of abstraction. It focusses on the system
environment and on  business and requirement specifications. 
At this level, the structure of the system  and its processing details are still unknown or undetermined.

\noindent 
-\textit{Platform Independent Models (PIM)}: this viewpoint focusses on the system functionality, hiding the details of any particular platform. 
%The specification defines those parts of the system that are independent of the development platform.

\noindent 
-\textit{Platform Specific Models (PSM)}: this viewpoint focusses on the functionality, in the context of a particular implementation platform.
Models at this level combine the platform-independent view with specific aspects of the target platform in order to implement the system.

@@

Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
They also assume that each service exports its interface.
So, the challenge of supporting non-functional properties is related to
\textit{(i)} the specification of the business rules of the application; and 
\textit{(ii)} dealing with the technical characteristics of the infrastructure where the application is  going to be executed.

>>>>

Model Driven Development (MDD)~\cite{Favre06arigorous} is a top-down approach proposed by the Object Management Group
(OMG)\footnote{\texttt{http://www.omg.org/mda}.} for designing and developing software systems. 
MDD provides a set of guidelines for  structuring  specifications by using \textit{models} to specify software systems at different levels of abstraction or \textit{viewpoints}:\\[1mm]
\noindent 
- \textit{Computation Independent Models (CIM)}: this viewpoint represents the software system at its highest level of abstraction. It focusses on the system
environment and on  business and requirement specifications. 
At this level, the structure of the system  and its processing details are still unknown or undetermined.

\noindent 
- \textit{Platform Independent Models (PIM)}: this viewpoint focusses on the system functionality, hiding the details of any particular platform. 
%The specification defines those parts of the system that are independent of the development platform.

\noindent 
- \textit{Platform Specific Models (PSM)}: this viewpoint focusses on the functionality, in the context of a particular implementation platform.
Models at this level combine the platform-independent view with specific aspects of the target platform in order to implement the system.

@@

Protocols and models implementing non-functional properties assume the existence of a global control of the artefacts implementing the application.
They also assume that each service exports its interface.
So, the challenge of supporting non-functional properties is related to
\textit{(i)} the specification of the business rules of the application; and 
\textit{(ii)} dealing with the technical characteristics of the infrastructure where the application is  going to be executed.

### 2013-10-22 SOFSEM2014/FinalVersion/main.tex
Final version to SOFSEM!

Works dealing with non-functional properties in service-oriented 
development can be organized in two main groups: those working on the \textit{modeling 
of particular non-functional properties or QoS attributes} and those 
proposing \textit{architectures or frameworks to manage and 
validate QoS attributes in web service composition processes}. 
The first group considers specific non-functional concerns (e.g., security)
which is modelled and then associated to functional models of the application. 
The work of Chollet et al.~\cite{CholletL09} defines a proposal to associate non-functional 
quality properties (security properties) to functional activities in a web service composition model. Schmeling et al.~\cite{SchmelingCM11} present 
an approach and also a toolset for specifying and implementing non-functional 
concerns in web service compositions.
Non-functional concerns are modelled and then related to a service composition 
represented in a BPMN diagram. 
Ovaska et al.~\cite{OvaskaEHPA10} present an approach to 
support quality management at design time. 
Quality requirements are modelled in a first phase and then
represented in an architectural model where quality requirements are associated to some
components of the model. 
We propose to model functional and non-functional properties at the same time during the software process.
We claim what this approach simplifies the web service development task. 
Non-functional properties are represented in our work as constraints and policies 
but more general quality model such as the proposed by~\cite{Goeb2011,Klass2009} 
could be taken in consideration in further works.

@@

Despite the variety of techniques proposed, there is not yet a consensus on a software methodology 
for web services. 
Some methodologies address the service-based 
development towards a standard or a new way to develop reliable applications. 
SOD-M and SOMF~\cite{somf} are MDD approaches for web services; 
S-Cube~\cite{scube2010book} is focused on the representation of business processes and 
service-based development; SOMA~\cite{soma} is a methodology 
for SOA solutions; 
%Extended SOA~\cite{PapazoglouH06} merges RUP and
%BPM~\cite{bpm} concepts for service modelling; 
DEVISE~\cite{DEVISE} is a
methodology for building service-based infrastructure for collaborative
enterprises. 
Other proposals include the WIED model~\cite{TongrungrojanaL04}, 
that acts as a bridge between business modeling and design models on one hand, and traditional 
approaches for software engineering applied to SOC on the other hand.

@@

\section*{Acknowledgements}
\small
This research is partly supported by the National Institute of Science and Technology for Software Engineering (INES\footnote{www.ines.org.br}), funded by CNPq (Brazil), grants 573964/2008-4 and 305619/2012-8; CAPES/UdelaR (Brazil/Uruguay) grant 021/ 2010; CAPES/STIC-AmSud (Brazil) grant 020/2010); MASAI project (TIN-2011-22617) financied by the Spanish Ministry of Science and Innovation and the Spanish Network on Service Science (TIN2011-15497-E) financed by the Spanish Ministry of Competitiveness and Economy. 
\normalsize

\bibliographystyle{splncs_srt}
\bibliography{biblio2} 
\end{document}

>>>>

Works dealing with non-functional properties in service-oriented 
development can be organized in two main groups: those working on the \textit{modelling 
of particular non-functional properties or QoS attributes} and those 
proposing \textit{architectures or frameworks to manage and 
validate QoS attributes in web service composition processes}. 
The first group considers specific non-functional concerns (e.g., security)
which is modelled and then associated to functional models of the application. 
The work of Chollet et al.~\cite{CholletL09} defines a proposal to associate non-functional 
quality properties (security properties) to functional activities in a web service composition model. Schmeling et al.~\cite{SchmelingCM11} present 
an approach and also a toolset for specifying and implementing non-functional 
concerns in web service compositions.
Non-functional concerns are modelled and then related to a service composition 
represented in a BPMN diagram. 
Ovaska et al.~\cite{OvaskaEHPA10} present an approach to 
support quality management at design time. 
Quality requirements are modelled in a first phase and then
represented in an architectural model where quality requirements are associated to some
components of the model. 
We propose to model functional and non-functional properties at the same time during the software process.
We claim what this approach simplifies the web service development task. 
Non-functional properties are represented in our work as constraints and policies 
but more general quality model such as the proposed by~\cite{Goeb2011,Klass2009} 
could be taken in consideration in further works.

@@

Despite the variety of techniques proposed, there is not yet a consensus on a software methodology 
for web services. 
Some methodologies address the service-based 
development towards a standard or a new way to develop reliable applications. 
SOD-M and SOMF~\cite{somf} are MDD approaches for web services; 
S-Cube~\cite{scube2010book} is focused on the representation of business processes and 
service-based development; SOMA~\cite{soma} is a methodology 
for SOA solutions; 
%Extended SOA~\cite{PapazoglouH06} merges RUP and
%BPM~\cite{bpm} concepts for service modelling; 
DEVISE~\cite{DEVISE} is a
methodology for building service-based infrastructure for collaborative
enterprises. 
Other proposals include the WIED model~\cite{TongrungrojanaL04}, 
that acts as a bridge between business modelling and design models on one hand, and traditional 
approaches for software engineering applied to SOC on the other hand.

@@

\section*{Acknowledgements}
\small
This research is partly supported by the National Institute of Science and Technology for Software Engineering (INES\footnote{www.ines.org.br}), funded by CNPq (Brazil), grants 573964/2008-4 and 305619/2012-8; CAPES/UdelaR (Brazil/Uruguay) grant 021/ 2010; CAPES/STIC-AmSud (Brazil) grant 020/2010); MASAI project (TIN-2011-22617) financed by the Spanish Ministry of Science and Innovation and the Spanish Network on Service Science (TIN2011-15497-E) financed by the Spanish Ministry of Competitiveness and Economy.

\normalsize  
\bibliographystyle{splncs_srt}
\bibliography{biblio2} 
\end{document}

### 2013-10-22 SOFSEM2014/FinalVersion/pisodm.tex
12 pages, Valeria's reccomndations.
Final version to SOFSEM!

%%%BEGIN{Text Included by Placido Neto for Camera-Ready}%%%%%%
$\pi$SOD-M environment is built on the top of Eclipse. 
We also used the Eclipse Modeling Framework (EMF) to define, edit and handle
(meta)-models. 
To automate the transformation models we use ATL~\cite{atl_manual} and Acceleo~\cite{acceleo}.

>>>>

%%%BEGIN{Text Included by Placido Neto for Camera-Ready}%%%%%%
$\pi$SOD-M environment is built on the top of Eclipse. 
We also used the Eclipse Modelling Framework (EMF) to define, edit and handle
(meta)-models. 
To automate the transformation models we use ATL~\cite{atl_manual} and Acceleo.
